Transition from forest to agriculture in the Brazilian Amazon from 1985 to 2021

### 2022-11-09 manuscript/manuscript.Rmd
Modify methods. Add text to results.

The estimations of transition from forest to agriculture were performed for the Amazon biome region, as defined by the Brazilian Institute of Geography and Statistics (IBGE), in 2019.
Transitions were calculated using land use and land cover classification data from MapBiomas [@Souza2020], which ranges from 1985 to 2021, at a spatial resolution of approximately 30 meters.

@@

3.  Calculate the first year of "Forest Formation" LULC class after the year calculated in step 2, for each pixel;

4.  Classify rows as "before" or "after" the occurrence of the year calculated in step 3, for each pixel;

5.  Calculate the last year of "Forest Formation" within the rows classified as "before", for each pixel;

@@

@@

We chose two ways to store and share the data.

-   One is to store the data from all three tables in three sets of Parquet files, in a structure of folders and sub folders;

-   The other is to convert the "trans_length.parquet" tables back to raster files.
    Raster files are useful to efficiently store spatial data, calculate areas, operate in relation to vector objects and perform data summaries, and it is also a very accessible format;

The new raster files are created based on the tiles downloaded from GEE (which are in the path "./data/raw_raster_tiles/"), using their extent values and its number of rows and columns (located on the "./data/trans_tabular_dataset/tiles_metadata.parquet" file).
For each tile file from "./data/raw_raster_tiles/", a set of new raster files are created based in the number of agriculture cycle ("agri_cycle") found in the tables, resulting in one raster for each agriculture cycle.
Each new raster file contains 6 bands:

-   Band 1 = "agri_code" (The agriculture class code that resulted from the transition);

-   Band 2 = "forest_year" (The last forest year before the transition, can be interpreted as the deforestation year);

-   Band 3 = "agri_year" (The first year of the agriculture establishment, being the end of the transition);

-   Band 4 = "trans_length" (The length in years of the transition from forest to agriculture);

-   Band 5 = "trans_cycle" (The number of the transition cycle in which the transition occurred);

-   Band 6 = "forest_type" (The type of the forest that suffered the transition);

The new raster files are saved in the path "./data/trans_raster_tiles/".

To facilitate the use of these new raster files, we also merged all the raster of each agriculture cycle into a single mosaic, which are saved in the path "./data/trans_raster_mosaic/".

@@

Although we named as deforestation the last years identified as forest before classification of antropic cover, we acknowledge that it is not a direct measurement of deforestation (such as PRODES), it is however a proxy to deforestation.

@@

```{r map-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Map of distribution of transitions from forests to agriculture in the Brazillian Amazon biome. The hexagonal cells represent the most common transition length, and do not reflect the amount of area of transitions inside a cell. Transitions are concentraded in the south (Mato Grosso state), and in the east (Pará and Maranhão states). The transition length ranges from 1 (blue tones) to 36 years (pink tones), and clusters of fast transitions (transitions closer to 1 year) can be discerned from clusters of slow transitions (transitions closer to 36 years). The histogram located in the bottom left shows that fast transitions are more common than slower transitions. The zoomed map in the bottom right shows the results in finer resolution, where it is possible to observe different transition lengths between properties."}

@@

```{r transbar-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Transition area per year and transition length. The bars represent the total amount of area at some state of the transition for each year. The color gradient in each bar represents the transition length related to a deforestation or a agriculture establishment event. Blue tones represents fast transitions (transitions closer to one year), pink tones represents slow transitions (transitions closer to 36 years). Transition events were separated by deforestation of primary forests (a) and secondary forests (c), and the subsequent  agriculture establishment of primary forests (b) and secondary forests (d). TODO: add labels for important years hwere public policies took place."}

@@

The deforestation area of primary forests increased largely from 1986 to 2003, which was followed be a steep decrease until 2008, when the deforested areas being converted to agriculture reached a stable rate.
From 1986 to approximately 1995, most of deforested areas suffered a slow transition, mostly were higher than 10 years, after this period, fast transitions started to become more common, specially from 2002 to 2004.

Agriculture establishment over areas of primary forests peaked in 2005 and 2013.
Despite similar rates between both years, their transition lengths differ greatly, in 2005 most of the transitions were faster than 10 years, while in 2013 the great majority of transitions were slower than 10 years.
The year of 2003 marked a change in the transition length of establishment of agriculture areas, after this year, most of the transitions happened in areas deforested at least 10 years before.
Even after the decrease of deforestation after 2002, agriculture areas are expanding over lands where deforestation happened before 2002.
However, after 2018, a sudden drop of agriculture establishment rate happened.

In 2006, the Brazilian Association of Vegetable Oil Industries (BIOVE) and the National Association of Cereal Exporters (ANEC) committed to avoid commercialization of soy grains harvested from areas deforested after 2006.
Our estimates of transitions show a decrease of deforested areas to be converted to agriculture after 2007, where it reached minimal values (Figure \@ref(fig:transbar-plot).a).
After 2006, agriculture establishment over deforested primary forests suffered a decrease, which stayed relatively stable until 2012, where a steep increase occurred, however, the new areas being occupied by agriculture were mainly over areas that were cleared more than a decade before (therefore, before 2006) (Figure \@ref(fig:transbar-plot).b) This shows that agriculture expansion did not halt after the soy moratorium, producers started expanding in old cleared areas.
Expansion of agriculture areas also expanded over cleared areas of secondary forests in 2012, but with an important amount of fast transitions.
The causes of the increase of agriculture establishment areas can be numerous, one main driver was the approval of a new Forest Code, in 2012, which is considered to have undermined the environmental protection of forests [@Kroger2017; @Pereira2019].

@@

```{r transridge-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Transition length patterns inside states that belongs to the Amazon biome."}

@@

The state of Amapá presented fast transitions along all the time series, where slow transitions are not as common.
In contrast, Acre shows a majority of slow transitions, in which only 2021 showed more fast transitions.There are three states where the pattern of transitions length across time are alike, Mato Grosso, Pará and Rondônia presented more fast transitions from 1995 to 2005, after this period, slow transitions became more common with time, until around 2015, where the occurrence of fast transitions were the majority again.

@@

```{r errorbar-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "TODO: Add title."}

@@

```{r errorscatter-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "TODO: Change line showing MAE to shaded areas showing quantiles of the errors. Add title."}

>>>>

The estimations of transition from forest to agriculture were performed for the Amazon biome region, as defined by the Brazilian Institute of Geography and Statistics (IBGE), in 2019.
Transitions were calculated using land use and land cover classification data from MapBiomas [@Souza2020], which ranges from 1985 to 2021, at a spatial resolution of approximately 30 meters.
The data was filtered to contain only pixels that were occupied by forest and agriculture at some period, and that are not considered as water, according to the Global Surface Water product provided by Copernicus [@Pekel2016].

@@

3.  Calculate the first year of "Forest Formation" LULC class after the year calculated in step 2;

4.  Classify rows as "before" or "after" the occurrence of the year calculated in step 3;

5.  Calculate the last year of "Forest Formation" within the rows classified as "before", and add 1 year to represent the deforestation year;

@@

The steps 2 to 7 are performed recursively to identify multiple transitions, in case they are present.
Because of the amount of data, the processing and storage was performed in batches of data, structured in raster tiles.

@@

After the calculation of transitions, the results are stored in three different types of tables, organized in a folder structure and stored as Apache Parquet files.

-   Tables that contains the spatial information (longitude, latitude) of each pixel, its unique id and the code of the municipality which contains the pixel.
    It is named as "mask_cells";

-   Tables with transition length values, the first and last year of the transition, the resulting agriculture type, and the number of the transition cycle.
    They also contain the unique id of each pixel (related to the table above);

-   Table with the LULC classes of all years within the transition, and also the first 5 years after the transition.

The three tables are related to each other and can be used altogether, and are separated by tiles.
Another table containing the metadata of each tile is also created, and holds the spatial characteristics of the tiles.
With this spatial information, it is possible to convert the tabular data back to spatial raster, with identical spatial properties as the MapBiomas classification data.

@@

Although we named as deforestation the last years identified as forest before classification of anthropic cover, we acknowledge that it is not a direct measurement of deforestation (such as PRODES), it is however a proxy to deforestation.

@@

```{r map-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Map of distribution of transitions from forests to agriculture in the Brazillian Amazon biome. The hexagonal cells represent the most common transition length, and do not reflect the amount of area of transitions inside a cell. Transitions are concentraded in the south (Mato Grosso state), and in the east (Pará and Maranhão states). The transition length ranges from 1 (blue tones) to 36 years (pink tones), and clusters of fast transitions (transitions closer to 1 year) can be discerned from clusters of slow transitions (transitions closer to 36 years). The histogram located in the bottom left shows that fast transitions are more common than slower transitions. The zoomed map in the bottom right shows the results in finer resolution, where it is possible to observe different transition lengths between properties."}

@@

```{r transbar-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Transition area per year and transition length. The bars represent the total amount of area at some state of the transition for each year. The color gradient in each bar represents the transition length related to a deforestation or a agriculture establishment event. Blue tones represents fast transitions (transitions closer to 0 years), pink tones represents slow transitions (transitions closer to 35 years). Transition events were separated by deforestation of primary forests (a) and secondary forests (c), and the subsequent  agriculture establishment of primary forests (b) and secondary forests (d)."}

@@

The deforestation area of primary forests increased largely from 1986 to 2003, which was followed be a steep decrease until 2009, when the deforested areas reached a stable rate.
From 1986 to approximately 1995, most of deforested areas suffered a slow transition, mostly were higher than 10 years, after this period, fast transitions started to become more common, specially from 2002 to 2004.

Agriculture establishment over areas of primary forests peaked in 2005 and 2013.
Despite similar rates between both years, their transition lengths differ greatly, in 2005 most of the transitions were faster than 10 years, while in 2013 the great majority of transitions were slower than 10 years.
The year of 2003 marked a change in the transition length of establishment of agriculture areas, after this year, most of the transitions happened in areas deforested at least 10 years before.
Even after the decrease of deforestation after 2002, agriculture areas are expanding over lands where deforestation happened before 2002.
However, after 2019, a sudden drop of agriculture establishment rate happened.

The causes of deforestation and agriculture establishment in the Brazilian Amazon are complex and diverse.
Political context, public policies, market prices and law enforcement can influence how these processes evolve over time.
In 2004, the Brazilian Government launched the Action Plan for the Prevention and Control of Deforestation in the Legal Amazon (PPCDAm), which was composed by many initiatives to curb deforestation [@West2021].
The PPCDAm was considered as a successful policy to slow deforestation rates in Brazil, with international recognition.
Our calculations from MapBiomas data reinforces the corelation of the PPCDAm with the reduction of deforestation after 2004, and reduction of the agriculture establishment after 2005.

In 2006, the Brazilian Association of Vegetable Oil Industries (BIOVE) and the National Association of Cereal Exporters (ANEC) committed to avoid commercialization of soy grains harvested from areas deforested after 2006.
Our estimates of transitions show a decrease of deforested areas to be converted to agriculture after 2008, where it reached minimal values (Figure \@ref(fig:transbar-plot).a).
After 2006, agriculture establishment over deforested primary forests suffered a decrease, which stayed relatively stable until 2012, where a steep increase occurred, however, the new areas being occupied by agriculture were mainly over areas that were cleared more than a decade before (therefore, before 2006) (Figure \@ref(fig:transbar-plot).b) This shows that agriculture expansion did not halt after the soy moratorium, producers started expanding in old cleared areas.
Expansion of agriculture areas also expanded over cleared areas of secondary forests in 2012, but with an important amount of fast transitions.
The causes of the increase of agriculture establishment areas can be numerous, one main driver was the approval of a new Forest Code, in 2012, which is considered to have undermined the environmental protection of forests [@Kroger2017; @Pereira2019].

@@

```{r transridge-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Transition length patterns inside states that belongs to the Amazon biome. Each year have a density estimate of the transition lengths, represented as colored curves. The peak of the curves represents transition length values with more frequency in one year of one state. Blue tones represents fast transitions (transitions closer to 0 years), pink tones represents slow transitions (transitions closer to 35 years)"}

@@

The state of Amapá presented fast transitions along all the time series, where slow transitions are not as common.
In contrast, Acre shows a majority of slow transitions, in which only 2021 showed more fast transitions.
There are three states where the pattern of transitions length across time are alike, Mato Grosso, Pará and Rondônia presented more fast transitions from 1995 to 2005, after this period, slow transitions became more common with time, until around 2015, where the occurrence of fast transitions were the majority again.

@@

```{r errorbar-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = " "}

@@

```{r errorscatter-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = " "}

### 2022-11-09 manuscript/manuscript.tex
Modify methods. Add text to results.

The estimations of transition from forest to agriculture were performed for the Amazon biome region, as defined by the Brazilian Institute of Geography and Statistics (IBGE), in 2019.
Transitions were calculated using land use and land cover classification data from MapBiomas \citep{Souza2020}, which ranges from 1985 to 2021, at a spatial resolution of approximately 30 meters.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Load raster and extract valid values into a table;
\item
  Calculate the year of first occurrence of a non-natural LULC class for each pixel;
\item
  Calculate the first year of ``Forest Formation'' LULC class after the year calculated in step 2, for each pixel;
\item
  Classify rows as ``before'' or ``after'' the occurrence of the year calculated in step 3, for each pixel;
\item
  Calculate the last year of ``Forest Formation'' within the rows classified as ``before'', for each pixel;
\item
  Calculate the first year of any agriculture type class within the rows classified as ``before'', for each pixel;
\item
  Calculate the difference between years from items 5 and 6 to get the LULC transition length in years, for each pixel;
\end{enumerate}

@@

We chose two ways to store and share the data.

\begin{itemize}
\item
  One is to store the data from all three tables in three sets of Parquet files, in a structure of folders and sub folders;
\item
  The other is to convert the ``trans\_length.parquet'' tables back to raster files.
  Raster files are useful to efficiently store spatial data, calculate areas, operate in relation to vector objects and perform data summaries, and it is also a very accessible format;
\end{itemize}

The new raster files are created based on the tiles downloaded from GEE (which are in the path ``./data/raw\_raster\_tiles/''), using their extent values and its number of rows and columns (located on the ``./data/trans\_tabular\_dataset/tiles\_metadata.parquet'' file).
For each tile file from ``./data/raw\_raster\_tiles/'', a set of new raster files are created based in the number of agriculture cycle (``agri\_cycle'') found in the tables, resulting in one raster for each agriculture cycle.
Each new raster file contains 6 bands:

\begin{itemize}
\item
  Band 1 = ``agri\_code'' (The agriculture class code that resulted from the transition);
\item
  Band 2 = ``forest\_year'' (The last forest year before the transition, can be interpreted as the deforestation year);
\item
  Band 3 = ``agri\_year'' (The first year of the agriculture establishment, being the end of the transition);
\item
  Band 4 = ``trans\_length'' (The length in years of the transition from forest to agriculture);
\item
  Band 5 = ``trans\_cycle'' (The number of the transition cycle in which the transition occurred);
\item
  Band 6 = ``forest\_type'' (The type of the forest that suffered the transition);
\end{itemize}

The new raster files are saved in the path ``./data/trans\_raster\_tiles/''.

To facilitate the use of these new raster files, we also merged all the raster of each agriculture cycle into a single mosaic, which are saved in the path ``./data/trans\_raster\_mosaic/''.

@@

\begin{figure}[h]
\includegraphics[width=17cm]{figs/map} \caption{Map of distribution of transitions from forests to agriculture in the Brazillian Amazon biome. The hexagonal cells represent the most common transition length, and do not reflect the amount of area of transitions inside a cell. Transitions are concentraded in the south (Mato Grosso state), and in the east (Pará and Maranhão states). The transition length ranges from 1 (blue tones) to 36 years (pink tones), and clusters of fast transitions (transitions closer to 1 year) can be discerned from clusters of slow transitions (transitions closer to 36 years). The histogram located in the bottom left shows that fast transitions are more common than slower transitions. The zoomed map in the bottom right shows the results in finer resolution, where it is possible to observe different transition lengths between properties.}\label{fig:map-plot}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=17cm]{figs/trans_length_cols} \caption{Transition area per year and transition length. The bars represent the total amount of area at some state of the transition for each year. The color gradient in each bar represents the transition length related to a deforestation or a agriculture establishment event. Blue tones represents fast transitions (transitions closer to one year), pink tones represents slow transitions (transitions closer to 36 years). Transition events were separated by deforestation of primary forests (a) and secondary forests (c), and the subsequent  agriculture establishment of primary forests (b) and secondary forests (d). TODO: add labels for important years hwere public policies took place.}\label{fig:transbar-plot}
\end{figure}

The deforestation area of primary forests increased largely from 1985 to 2002, which was followed be a steep decrease until 2007, when the deforested areas being converted to agriculture reached a stable rate.
From 1985 to approximately 1995, most of deforested areas suffered a slow transition, mostly were higher than 10 years, after this period, fast transitions started to become more common, specially from 2001 to 2003.

Agriculture establishment over areas of primary forests peaked in 2005 and 2013.
Despite similar rates between both years, their transition lengths differ greatly, in 2005 most of the transitions were faster than 10 years, while in 2013 the great majority of transitions were slower than 10 years.
The year of 2003 marked a change in the transition length of establishment of agriculture areas, after this year, most of the transitions happened in areas deforested at least 10 years before.
Even after the decrease of deforestation after 2002, agriculture areas are expanding over lands where deforestation happened before 2002.
However, after 2018, a sudden drop of agriculture establishment rate happened.

In 2006, the Brazilian Association of Vegetable Oil Industries (BIOVE) and the National Association of Cereal Exporters (ANEC) committed to avoid commercialization of soy grains harvested from areas deforested after 2006.
Our estimates of transitions show a decrease of deforested areas to be converted to agriculture after 2007, where it reached minimal values (Figure \ref{fig:transbar-plot}.a).
After 2006, agriculture establishment over deforested primary forests suffered a decrease, which stayed relatively stable until 2012, where a steep increase occurred, however, the new areas being occupied by agriculture were mainly over areas that were cleared more than a decade before (therefore, before 2006) (Figure \ref{fig:transbar-plot}.b) This shows that agriculture expansion did not halt after the soy moratorium, producers started expanding in old cleared areas.
Expansion of agriculture areas also expanded over cleared areas of secondary forests in 2012, but with an important amount of fast transitions.
The causes of the increase of agriculture establishment areas can be numerous, one main driver was the approval of a new Forest Code, in 2012, which is considered to have undermined the environmental protection of forests \citep{Kroger2017, Pereira2019}.

@@

\begin{figure}[h]
\includegraphics[width=17cm]{figs/trans_ridge} \caption{Transition length patterns inside states that belongs to the Amazon biome.}\label{fig:transridge-plot}
\end{figure}

The state of Amapá presented fast transitions along all the time series, where slow transitions are not as common.
In contrast, Acre shows a majority of slow transitions, in which only 2021 showed more fast transitions.There are three states where the pattern of transitions length across time are alike, Mato Grosso, Pará and Rondônia presented more fast transitions from 1995 to 2005, after this period, slow transitions became more common with time, until around 2015, where the occurrence of fast transitions were the majority again.

@@

\begin{figure}[ht]
\includegraphics[width=17cm]{figs/error_bars} \caption{TODO: Add title.}\label{fig:errorbar-plot}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=17cm]{figs/error_scatter} \caption{TODO: Change line showing MAE to shaded areas showing quantiles of the errors. Add title.}\label{fig:errorscatter-plot}
\end{figure}

>>>>

The estimations of transition from forest to agriculture were performed for the Amazon biome region, as defined by the Brazilian Institute of Geography and Statistics (IBGE), in 2019.
Transitions were calculated using land use and land cover classification data from MapBiomas \citep{Souza2020}, which ranges from 1985 to 2021, at a spatial resolution of approximately 30 meters.
The data was filtered to contain only pixels that were occupied by forest and agriculture at some period, and that are not considered as water, according to the Global Surface Water product provided by Copernicus \citep{Pekel2016}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Load raster and extract valid values into a table;
\item
  Calculate the year of first occurrence of a non-natural LULC class for each pixel;
\item
  Calculate the first year of ``Forest Formation'' LULC class after the year calculated in step 2;
\item
  Classify rows as ``before'' or ``after'' the occurrence of the year calculated in step 3;
\item
  Calculate the last year of ``Forest Formation'' within the rows classified as ``before'', and add 1 year to represent the deforestation year;
\item
  Calculate the first year of any agriculture type class within the rows classified as ``before'', for each pixel;
\item
  Calculate the difference between years from items 5 and 6 to get the LULC transition length in years, for each pixel;
\end{enumerate}

The steps 2 to 7 are performed recursively to identify multiple transitions, in case they are present.
Because of the amount of data, the processing and storage was performed in batches of data, structured in raster tiles.

@@

After the calculation of transitions, the results are stored in three different types of tables, organized in a folder structure and stored as Apache Parquet files.

\begin{itemize}
\item
  Tables that contains the spatial information (longitude, latitude) of each pixel, its unique id and the code of the municipality which contains the pixel.
  It is named as ``mask\_cells'';
\item
  Tables with transition length values, the first and last year of the transition, the resulting agriculture type, and the number of the transition cycle.
  They also contain the unique id of each pixel (related to the table above);
\item
  Table with the LULC classes of all years within the transition, and also the first 5 years after the transition.
\end{itemize}

The three tables are related to each other and can be used altogether, and are separated by tiles.
Another table containing the metadata of each tile is also created, and holds the spatial characteristics of the tiles.
With this spatial information, it is possible to convert the tabular data back to spatial raster, with identical spatial properties as the MapBiomas classification data.

@@

\begin{figure}[ht]
\includegraphics[width=17cm]{figs/map} \caption{Map of distribution of transitions from forests to agriculture in the Brazillian Amazon biome. The hexagonal cells represent the most common transition length, and do not reflect the amount of area of transitions inside a cell. Transitions are concentraded in the south (Mato Grosso state), and in the east (Pará and Maranhão states). The transition length ranges from 1 (blue tones) to 36 years (pink tones), and clusters of fast transitions (transitions closer to 1 year) can be discerned from clusters of slow transitions (transitions closer to 36 years). The histogram located in the bottom left shows that fast transitions are more common than slower transitions. The zoomed map in the bottom right shows the results in finer resolution, where it is possible to observe different transition lengths between properties.}\label{fig:map-plot}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=17cm]{figs/trans_length_cols} \caption{Transition area per year and transition length. The bars represent the total amount of area at some state of the transition for each year. The color gradient in each bar represents the transition length related to a deforestation or a agriculture establishment event. Blue tones represents fast transitions (transitions closer to 0 years), pink tones represents slow transitions (transitions closer to 35 years). Transition events were separated by deforestation of primary forests (a) and secondary forests (c), and the subsequent  agriculture establishment of primary forests (b) and secondary forests (d).}\label{fig:transbar-plot}
\end{figure}

The deforestation area of primary forests increased largely from 1986 to 2003, which was followed be a steep decrease until 2009, when the deforested areas reached a stable rate.
From 1986 to approximately 1995, most of deforested areas suffered a slow transition, mostly were higher than 10 years, after this period, fast transitions started to become more common, specially from 2002 to 2004.

Agriculture establishment over areas of primary forests peaked in 2005 and 2013.
Despite similar rates between both years, their transition lengths differ greatly, in 2005 most of the transitions were faster than 10 years, while in 2013 the great majority of transitions were slower than 10 years.
The year of 2003 marked a change in the transition length of establishment of agriculture areas, after this year, most of the transitions happened in areas deforested at least 10 years before.
Even after the decrease of deforestation after 2002, agriculture areas are expanding over lands where deforestation happened before 2002.
However, after 2019, a sudden drop of agriculture establishment rate happened.

The causes of deforestation and agriculture establishment in the Brazilian Amazon are complex and diverse.
Political context, public policies, market prices and law enforcement can influence how these processes evolve over time.
In 2004, the Brazilian Government launched the Action Plan for the Prevention and Control of Deforestation in the Legal Amazon (PPCDAm), which was composed by many initiatives to curb deforestation \citep{West2021}.
The PPCDAm was considered as a successful policy to slow deforestation rates in Brazil, with international recognition.
Our calculations from MapBiomas data reinforces the corelation of the PPCDAm with the reduction of deforestation after 2004, and reduction of the agriculture establishment after 2005.

In 2006, the Brazilian Association of Vegetable Oil Industries (BIOVE) and the National Association of Cereal Exporters (ANEC) committed to avoid commercialization of soy grains harvested from areas deforested after 2006.
Our estimates of transitions show a decrease of deforested areas to be converted to agriculture after 2008, where it reached minimal values (Figure \ref{fig:transbar-plot}.a).
After 2006, agriculture establishment over deforested primary forests suffered a decrease, which stayed relatively stable until 2012, where a steep increase occurred, however, the new areas being occupied by agriculture were mainly over areas that were cleared more than a decade before (therefore, before 2006) (Figure \ref{fig:transbar-plot}.b) This shows that agriculture expansion did not halt after the soy moratorium, producers started expanding in old cleared areas.
Expansion of agriculture areas also expanded over cleared areas of secondary forests in 2012, but with an important amount of fast transitions.
The causes of the increase of agriculture establishment areas can be numerous, one main driver was the approval of a new Forest Code, in 2012, which is considered to have undermined the environmental protection of forests \citep{Kroger2017, Pereira2019}.

@@

\begin{figure}[ht]
\includegraphics[width=17cm]{figs/trans_ridge} \caption{Transition length patterns inside states that belongs to the Amazon biome. Each year have a density estimate of the transition lengths, represented as colored curves. The peak of the curves represents transition length values with more frequency in one year of one state. Blue tones represents fast transitions (transitions closer to 0 years), pink tones represents slow transitions (transitions closer to 35 years)}\label{fig:transridge-plot}
\end{figure}

The state of Amapá presented fast transitions along all the time series, where slow transitions are not as common.
In contrast, Acre shows a majority of slow transitions, in which only 2021 showed more fast transitions.
There are three states where the pattern of transitions length across time are alike, Mato Grosso, Pará and Rondônia presented more fast transitions from 1995 to 2005, after this period, slow transitions became more common with time, until around 2015, where the occurrence of fast transitions were the majority again.

@@

\begin{figure}[ht]
\includegraphics[width=17cm]{figs/error_bars} \caption{ }\label{fig:errorbar-plot}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=17cm]{figs/error_scatter} \caption{ }\label{fig:errorscatter-plot}
\end{figure}
